### 2021-05-07 Paper/lit_review.tex
big push in latex
the final push :)

This section will be broke into two components. The first section will adress  theoretical models for asset purchasing and then we will introduce some of the mythology for  EGS factor in the asset pricing framework.\\

 \hl{add a section on the metholoogy for the 2 pass sort use} \cite{Kleibergen2013UnexplainedFA} and \cite{Fama1973RiskRA} \\
The modern foundation of asset pricing began with the capital asset pricing model based on Sharp's 1964's work which used a regression analysis to relate market returns to individual security returns \cite{Sharpe1964CAPITALAP}. The CAPM model was later iterated on by Fama french in 1993 with the 3 factor, which considered several other terms to predict stock returns. Fama and French added size and book to market terms to the CAPM model. They found that these three coefficients created a more explanatory model which demonstrated the relationship that value companies(low book to market) and small companies experience larger excess returns \cite{Fama1992TheCO}. Fama and French again innovated on the three-factor model in 2015 by adding two factors to the regression. In this model, Fama and French consider Investment and profitability while holding the other regression stable factors. The five-factor model analysis found that the addition of the investment and profitability factor made the value factor redundant \cite{Fama2013AFA}. The five-factor framework analysis has been recreated on international exchanges to test the effects on foreign markets. In recreation by Fama French and another study by Zhong and Li, the general relationships demonstrated in the five-factor model are also applicable in China, Japan, and Australian markets \cite{Fama2015InternationalTO, LIN2017141, Chiah2016A}. However, when the Fama French regression was applied to the Australian equity market, the authors did not find the significance of the value factor to subside like was visible in the 2015 paper \cite{Chiah2016A}.  In addition to expanding the five-factor analysis geographically, some scholars have expanded the five-factor model by adding additional terms to assess momentum effects.  In a paper published by  Dirkx and  Peter, the authors expanded the five-factor model by applying the Fama French framework to the German market while also adding a sixth term to assess the effect of risk premium with the CDAX constituents' momentum. Another study by Gregory and Stead used the sixth factor as an ESG factor to analyze if the ESG effect played a significant role in determining risk premia. This paper found that  "the risk premium of the sustainability factor is positive and significant\cite{Gregory2020TheGP}."

In addition to Gregory and Stead's paper, several other authors have used the Fama French model to address the question concerning ESG and risk premia. To set up an ESG Fama French model, many academic journal articles use an index as a positive ESG portfolio and another index as a negative portfolio to create the ESG factor. In Gregory and Stead's paper, the author uses the S\&P ESG index as the high ESG index and the S\&P petroleum index as the low index to create a factor by subtracting the returns from the high ESG portfolio from the low ESG portfolio \cite{Gregory2020TheGP}. In Dorfleitner et al.'s paper, the authors collect Thomas Reuter's controversies and combined ESG score for 2500 companies. With the ranked ESG score data, the author then creates two portfolios by separating the top decile score and the bottom deciles scores, which act as the ESG and contra ESG portfolios. Similar to Gregory et al. paper, the author subtract the top 10\% from the bottom 10\% to create the ESG factor \cite{Dorfleitner2020ESGCA}.

>>>>

The modern foundation of asset pricing began with the capital asset pricing model based on Sharp's 1964 work which used a regression analysis to relate market returns to individual security returns \cite{Sharpe1964CAPITALAP}. The CAPM model was later iterated on by Fama french in 1993 with the 3 factors, which considered size,  book, and the market terms from the CAPM model. Fama French found that these three coefficients created a more explanatory model which demonstrated the relationship that value companies (low book to market) and small companies experience larger excess returns \cite{Fama1992TheCO}. In addition to adding the two factors, Fama French also incorporated the two-pass sort methodology, which isolates portfolio sensitivity and risk premia in a multi sep regression process \cite{Kleibergen2013UnexplainedFA, Fama1973RiskRA}.

Fama and French innovated on the three-factor model in 2015 by adding two additional regressors, investment and profitability. The five-factor model analysis found that the addition of the investment and profitability factor made the value factor redundant \cite{Fama2013AFA}. The reliability of the five-factor has been shown by replications on international exchanges and has been found to hold true in Germany, China, and Japan but did not hold true for Australia \cite{Fama2015InternationalTO, LIN2017141, Chiah2016A}. The Fama French five factor model is the standard in asset pricing and is the basis for many academic research projects that extend the five-factor model to address additional unpriced factors \cite{Dirkx2020TheFF}. Gregory and Stead use a sixth factor as an ESG factor to analyze if the ESG effect played a significant role in determining risk premia. This paper found that  ``the risk premium of the sustainability factor is positive and significant"\cite{Gregory2020TheGP}.

Literature on ESG investing has become a quickly growing academic interest, with researchers finding both strong evidence for a non-priced ESG factor and no evidence of a noticeable ESG factor. In addition to Gregory and Stead's paper, several other authors have used the Fama French model to address the question concerning ESG and risk premia. To set up an ESG Fama French model, many academic journal articles use an index as a positive ESG portfolio and another index as a negative portfolio to create the ESG factor. In Gregory and Stead's paper, the author uses the S\&P ESG index as the high ESG index and the S\&P petroleum index as the low index to create a factor by subtracting the returns from the high ESG portfolio from the low ESG portfolio \cite{Gregory2020TheGP}. In Dorfleitner et al.'s paper, the authors collect Thomas Reuter's controversies and combined ESG score for 2500 companies. With the ranked ESG score data, the author then creates two portfolios by separating the top decile score and the bottom deciles scores, which act as the ESG and contra ESG portfolios. The author finds that there are measurable excess returns and contributes. The main findings for this paper is the realization that small companies, particularly those who stay out of controversy, trend to be undervalued and appreciate compared to similar size companies with lots of controversy \cite{Dorfleitner2020ESGCA}.

Although authors have found positive evidence, other authors have found no such evidence of a relationship. Work completed by Liou investigates the MSCI ESG rating data to construct equal-weighted portfolios as far back as 1991  and performs a Fama French analysis. In the paper, the author finds that in a post-GFC test, there is no excess return for a long strong ESG short poor ESG factor. However, the author does note that there is a relationship to the size and the coefficient for the ESG factor \cite{Lioui2018ESGFI}.

Additionally, some authors have looked at the excess returns on ESG portfolios and assess the effects over calendar years. Work from Bennani et al. explored the post-GFC landscape for ESG investing and found that in North American stocks from 2010-2013, investing with ESG screening reduced excess returns; however, from 2014-2017, the authors found that ESG screening created 3.3\% excess returns annually \cite{Bennani2018HowEI}.
%. These results strongly indicate that this is, on the one hand, driven by low-rated smaller companies (“small sinners”) and clean-coated frms with regard to controversies (“silent saints”) on the other hand.

%Among others, Derwall et al. (2005) and Edmans (2011), who link the doing good while doing well-hypothesis with the managerial myopia theory, conclude that short-term investors are unable (or unwilling) to price the long-term benefts of those activities correctly and therefore undervalue stocks of companies with high levels of engagement in environmental or social aspects, leading to higher returns in the long-run for the respective stocks when compared with other stocks.
%with the fndings of Dorfeitner et al. (2018), who conclude that the benefts of socially responsible activities (measured by the abnormal stock returns) are produced by unexpected additional cash fows which occur mid-to-long term.
%Furthermore, from an investor’s standpoint, having a “clean coat” with regard to controversies is especially proftable for smaller companies, as the absence of these scandals may be overlooked and incorrectly incorporated in the market prices. Thus, one might say that the respective companies “fy under the radar”.

### 2021-05-07 Paper/motivation.tex
the final push :)

Asset pricing has a long history of models and theories governing how investors make decisions. Over the last decade, investors have become increasingly interested in factors outside of traditional financial metrics when assessing the possibility of an investment's returns. Much of this enthusiasm has been based on an ESG, Environmental Social Governance, framework. ESG investing considers how firms respond to these factors as potential impetuses for excess performance. This increased attention is especially visible in Morgan Stanley's 2019 survey of the ESG landscape. Morgan Stanley  found that 75\% of assessment managers say that "their firm has adopted sustainable investing." Only two years prior, in 2016, just 10\% of surveyed asset managed had adopted sustainable investing, demonstrating the rapid speed of adoption for ESG investing analysis \cite{morgan_stanley_sustainable_2019}. Practitioners have utilized the ESG framework to reduce risk in their investments while also finding abnormal returns.

Global climate change presents a host of possible and sizable financial risks from increased water levels, which could threaten major global cities' safety, including Miami, Honk Kong, and Shanghai, with the potential to displace 100's of millions of people \cite{holder_three-degree_nodate}. Global climate change is also shifting the landscape for global agricultural production as more extreme weather conditions, including droughts, pose problems to the most fundamental building block of economic activity. In addition to risk mitigation, investors look to utilize ESG principle for superior returns. A research report published by McKinsey and Company linked gender diversity on executive teams to higher profitability and superior value creation \cite{mckinsey_diversity}.

In addition to risk investing professionals, ESG investing is appealing to a younger set of investors interested in a plethora of global issues from global climate change to diversity and inclusion in the workplace. A study conducted by Morgan Stanley found that millennials investors were twice as likely as the general investing population to invest in a company with social or environmental goals.

This new wave of investor sentiment has created strong tailwinds for sustainable investment projects. In 2020 alone, fund flows into sustainable funds grew by 51 billion dollars which is up over 100\% for the 2019 fund flow into sustainable investment products of about 21, placing total assets under management in sustainable products at over 250B. Fund flows into sustainable products for 2020 accounted for a little less a quarter of all US mutual fund flow and about one-fifteenth of all ETF fund flow according to morningstar \cite{monring_star_ESG}.

To address the impact of ESG factors on stock preformance, this paper is broken into five sections. \textit{Section \ref{Literature}} will provide a  brief literature review of papers looking at abnormal performance around ESG criteria as well as discussion of asset pricing models to capture ESG factors in performance. Next \textit{ Section \ref{Data}}  will provide a brief discussion of the data used in this paper as well as the data cleaning process and summary statistics. Following the Data, \textit{Section \ref{Methodology}} will breakdown of the methodology for the paper. Finally, \textit{Section \ref{Results}}  will summarize the results of the regressions, \textit{Section \ref{Robustness}} will check the robustness of the results and \textit{Section \ref{Conclusion}} will provide a conclusion and possible extensions for next steps.

>>>>

Over the last decade, investors have become increasingly interested in monitoring factors beyond traditional financial metrics to find opportunities to achieve excess returns. In this pursuit,  many investors have focused their attention on the ESG, Environmental Social Governance, framework. Using the ESG framework, investors evaluate a firm's sustainability and societal impact, which are typically overlooked in traditional asset pricing models. This method of investing has garnered quick traction as global climate change threatens the longevity of critical global infustructure, and diversity initiatives shape the modern workforce  \cite{holder_three-degree_nodate, mckinsey_diversity}. A study conducted by Morgan Stanley found in 2016, just 10\% of surveyed asset managers had adopted sustainable investing, a number that quickly grew to over 70\% by 2018 \cite{morgan_stanley_sustainable_2019}.

This new wave of investor sentiment has created strong tailwinds for sustainable investment products. In 2020 alone, fund flows into sustainable funds grew by \$51 billion  which is up over 100\% from the 2019 fund flow into sustainable investment products of about \$21 billion, placing total assets under management in sustainable products at over \$250 billion. Fund flows into sustainable products for 2020 accounted for a little less a quarter of all US mutual fund flow and about one-fifteenth of all ETF fund flow according to Morningstar \cite{monring_star_ESG}.

In this paper, I assess the impact of ESG investing on the excess returns through a five-factor Fama French framework. I establish two different methods for defining ESG risk premia and find dissimilar results. When defining ESG risk premia as the difference between the MSCI ESG index and the S\&P petroleum index, I find a robust risk premia, which suggests that a one-unit change in beta results in an 8.7\% increase in excess returns for size decile portfolios. However, when defining ESG risk premia as the difference between high ESG firms minus low ESG firms as rated by MSCI, there are no statistically measurable risk premia. Further, when analyzing the risk premia at the one-digit SIC level, only one SIC code yields a statistical relationship between excess returns and the ESG factor.

### 2021-05-07 Paper/robustness.tex
the final push :)

As discussed in section \ref{Results}, additional tests are required to ensure the methodology's robustness. The first robustness check I will run is a repetition of the two-pass sort for where the ESG factor is calculated asMSCI ESG index minus the petroleum index (EMP). Table 6 confirms the statistical significance of the market term, the SMB term, and the CMA term. All three of these terms were found to be statistically significant in the initial regression as well. Beyond the coefficients,  the results in table 6 demonstrate a slightly lower $R^2$ for both the Beta decile and the size decile variants, demonstrating that the EMP term contributes to the explanation of excess returns. In addition, the constant term for both size and beta decile portfolios increased in magnitude as a result of omitting the EMP. As the coefficient term gets closer to zero, the factors do a better job describing the returns and indicate that there is no excess return that is not explained. 
 %as dicussed in section \ref{Data} [talk about using mining instead of petrolum ]

 \begin{center}
    \paperspacingnarrow
    \input{Code/data/FF_pass_2_no_ESG}\\
    \textbf{Table 6:} second pass results for the size and beta decile portfolios without ESG factor\\
    \paperspacingwide
\end{center}
Another robustness check that is essential to consider is the time-based effects on the model. Although we have demonstrated that the EMP factor is statistically significant and explains part of our model, it is essential to understand if this is ubiquitous across time in the sample or if the  ESG factor has shifted in risk premia over time. To address this robustness concern, I have repeated the two-pass sort with the size decile portfolio by adding a time-based dummy variable. Table 7 details the results from the regressions. In 2015 2016, 2018, and 2019 there is statistical significance for the EMP term and a positive association between EMP and excess returns. The ESG factor showed the most strength in 2019 as well as the greatest coefficient value. In 2019 a one-unit change in beta would result in a 25.9\% increase in excess returns. The coefficient values in 2017 appeared to hold a negative relationship as a one-unit change in ESG beta would have resulted in a reduction in excess returns by 10.5\%. Although in 2020, there was a positive relationship, the coefficient is not statistically significant. However, in 2020, the CMA factor, which has been statistically significant at the 1\% level for all other tests, also does not appear to be statistically significant. 2020 represents a bit of a deviation due to the larger market condition from the COVID-19 pandemic rather than the EMP factor's long-run behavior.

\begin{center}
    \paperspacingnarrow
    \input{Code/data/Time_FF_pass_2}\\
    \textbf{Table 7:} Shows the results for the second pass in a two pass sort. The two pass sort used size based deciles for the first pass. Columns (1)-(6) represent individual years of data. 
    \paperspacingwide
\end{center}
 \textbf{MSCI data 2018-2020}\\
 \begin{center}
    \paperspacingnarrow
\input{Code/data/All_Industry_FF_pass_2_past_2018}\\
\textbf{Table 8:} SIC top 10\% minus bottom 10\% 2018-2020
    \paperspacingwide
\end{center}
 \textbf{MSCI SIC level analysis}\\

  \begin{center}
    \paperspacingnarrow
    \input{Code/data/Industry_FF_pass_2}\\
    \textbf{Table 6:} second pass results for the size and beta decile portfolios without ESG factor\\
    \paperspacingwide
\end{center}

 WE ADJUSTED THE YEAR to 2019!!
 problems we can never solve for both of these indexes does not do a great job capturing small companies so size might be a bit of a probelm overall as the small companies do not have ther ESG scores baked in as MSCI does not have the resources to rank every company

 when doing the SIC porlfios did not have data for SIC starting with 5 b/c they were all BB companies

dicuss limitation of only one set of ESG factors for MSCI

 we used the Size deciles on a go froward basis when analysis the data as the statistical relationship was more apparent

>>>>

As discussed in section \ref{Results}, additional tests are required to ensure the robustness of the methodology. The first robustness check I will run is a repetition of the two-pass sort for where the ESG factor is calculated as MSCI ESG index minus the petroleum index (EMP). Table 6 confirms the statistical significance of the market term, the SMB term, and the CMA term. Beyond the coefficients,  the results in table 6 demonstrate a slightly lower $R^2$ for both the Beta decile and the size decile variants, demonstrating that the EMP term contributes to the explanation of excess returns. In addition, the constant term for both size and beta decile portfolios increased in magnitude as a result of omitting the EMP. As the coefficient term gets closer to zero, the factors do a better job describing the excess return and relationship and eliminates all other sources of unexplained excess returns.

 \begin{center}
    \paperspacingnarrow
    \input{Code/data/FF_pass_2_no_ESG}\\
    \textbf{Table 6:} shows  second pass results in a teo pass sort for the size and beta decile portfolios without EMP factor\\
    \paperspacingwide
\end{center}
Another robustness check that is essential to consider is the time-based effects on the model. Although I have demonstrated that the EMP factor is statistically significant and explains part of our model, it is essential to understand if this is ubiquitous across time in the sample or if the  ESG factor has shifted in risk premia over time. To address this robustness concern, I have repeated the two-pass sort with the size decile portfolio by adding a time-based dummy variable. Table 7 details the results from this regression. In 2015 2016, 2018, and 2019 there is statistical significance for the EMP term and a positive association between EMP and excess returns. The ESG factor showed the most statistical strength in 2019 with the greatest excess return. In 2019 a one-unit change in beta would result in a 25.9\% increase in excess returns. These results are consistent with Bennani et al.'s paper decomposing risk premia over time for 2015 and 2016. The coefficient value of the EMP term in 2017  holds a negative relationship as a one-unit change in ESG beta resulted in a reduction in excess returns by 10.5\%. In 2020, there was a positive relationship, however the coefficient is not statistically significant. The CMA factor, which has been statistically significant at the 1\% level for all other tests, also does not appear to be statistically significant for 2020. The 2020 calendar year represents a bit of a deviation from typical trends likely due to the larger market condition from the COVID-19 pandemic rather than the EMP factor's long-run behavior. In addition to this set of robustness checks, it would be beneficial to repeat the series of tests replacing the S\&P petroleum index with the S\&P mining index as the contra ESG portfolio. Both of the proposed indexes fall under the first digit  SIC code 1  and are exposed to similar ESG risks; however, the additional test can isolate risk premia and test if the factor is capturing petroleum industry effect or a robust ESG factor.

\begin{center}
    \paperspacingnarrow
    \input{Code/data/Time_FF_pass_2}\\
    \textbf{Table 7:} shows the results for the second pass in a two pass sort. The two pass sort used size based deciles for the first pass. Columns (1)-(6) represent individual years of data from 2015-2020. 
    \paperspacingwide
\end{center}

 Moving on to the second set of robustness checks with a slightly different definition of the ESG factor.  As described in the methodology and the results, I created another definition for measuring the effect of ESG by subtracting the Best 10\% of frim from the bottom 10\% of firms as defined by MSCI ESG scores. As demonstrated in the results section, there was no measured statistical relationship between this factor and the excess returns. To further break down this relationship, I decomposed the excess returns on a first digit SIC level to assess if some industries are more sensitive to ESG factors than others. Some industries are likely to be inherently ESG unfriendly and might not be substantially affected by an ESG factor.

To create this analysis, I repeated the two-pass sort methodology from the results using the size decile portfolios and isolated the returns for each of the first digit SIC code industries.  I regressed the same ESG factor described in the results on this subset to assess the different effects the factor has on each industry. As the MSCI ESG rating dataset for this analysis did not have robust data for industries with first digit SIC codes 0, 5, or 9, I elected not to analyze these SIC industries as the construction of the factor did not account for them.

Table 8 demonstrates the results from this new set of regressions. For all of the available  SIC portfolios, the market term and the SMB term are highly statistically significant at the 1\% level, which is consistent with the results from both the EMP and the initial MSCI ESG rating analysis. The ESG factor response to each industry is not uniform. Industries with first digit SIC codes 3,4,6,7, and 8 have small non statistically significant coefficients.  The first digit SIC code 2 demonstrated a statistically significant negative relationship to the ESG factor, demonstrating that the high ESG beta portfolio would harm excess returns. However, the first digit SIC code 1 demonstrated a highly statistically significant positive coefficient.  A one-unit increase in ESG beta would result in an 11.6\% increase in excess returns for first digit SIC industry 1. These results confirm results from part 1 of the analyses with the EMP factor as the SIC code 1 encompasses the petroleum industry. Future tests should be constructed to ensure that the ESG factor is picking up on more than negative excess returns from investing in the petroleum industry.

\begin{center}
    \paperspacingnarrow
    \input{Code/data/NEW_Industry_FF_pass_2}\\
    \textbf{Table 8:} shows the results for the second regression results in a two-pass sort. Columns (1)-(7) represent the first digit SIC code industries and their respective risk premia to the Fama French five-factor as well as the ESG factor. SIC code 1 covers the mining and construction industry, SIC codes 2-3 covers manufacturing industry, SIC code 4 covers transportation, communications, electric, gas, and sanitary service SIC code 6 covers finance, insurance, and real estate SIC code 7-8 covers services\\

    \paperspacingwide
\end{center}

A final robustness check is conducted to address another concern about the MSCI ESG data. As explained in the data section, the MSCI ESG data used to create the factor is static. That is to say, the data for ESG scores is only a snapshot of ESG ratings for April 2021. The use of the entire window from June 2011 to March 2021 to assess excess returns might be a bad characterization of the ESG landscape as a company that could have been a low ESG company might have switched to a high ESG company in that time. To address this concern, I recreated the regression from table 5 by limiting the time horizon to 2018-2021. Although there could have been some movement in the ESG scores during the modified time frame, a companies ability to respond to ESG objectives has a degree of stickiness as corporate initiatives often take several years to develop and implement, leading to a lag effect.  Looking at the results from table 9, the market, SMB, and CMA factors are all statistically significant, which was also observed in table 5. The ESG factor in table 9, similar to table 5, is not statistically significant for either the beta or size decile portfolios. The magnitude of the  ESG factor for the size-based portfolio has increased from .004 to .013, which indicates the EGS factor might have a larger impact on excess returns. However, the beta decile portfolio has changed signs from .0145 to -.013, which indicates that the EGS factor has a negative impact on excess returns. The muddled resulted mixed with the statistical insignificance do not provide evidence of a shorter time horizon improving the accuracy of the ESG factor as a predictor of stock performance. An additional source of error could be a result of the sample bias associated with the companies MSCI has rated. MSCI's publicly available ESG ratings data has 545 US firms. Of the rated companies, 439 (or 99\% of rated firms) have a market cap greater than \$1 billion. As a result of the large market cap bias, the ESG factor might not accurately reflect the risk premia associated with all ESG investing.

 \begin{center}
    \paperspacingnarrow
    \input{Code/data/All_Industry_FF_pass_2_past_2018}\\
    \textbf{Table 9:} shows the same regression from table 5 with an altered time frame to only include data from 2018-2021
    \paperspacingwide
\end{center}
